A Bayesian ranking statistic to find high-mass black holes in LIGO--Virgo data

### 2021-03-13 main.tex
Updates from Overleaf

The most promising avenues to find $100−1000\ \msun$ black holes are either to identify stellar binaries, which are expected to be common or to detect them in gravitational radiation. Ground-based gravitational wave limits on the low end of this mass regime are continuing to improve.

@@

Since the 1970s, there has been an accumulation of evidence for stellar-mass and supermassive black holes. In April 2019, the Event Horizon Telescope provided the first visual evidence of the supermassive black hole M87~\cite{m87photo}. As of January 2021, the LIGO Scientific Collaboration has confirmed $\sim50$ binary black hole systems and listed numerous candidate events~\cite{GWTC1, GWTC2, gracedb}. Additionally, since the public release of LIGO's first and second observing run's data (O1 and O2), search pipelines such as \pycbc~\cite{pycbc_code} and a research team at the Institute for Advanced Study (IAS)~\cite{IAS0, IAS1, IAS2}, have independently identified additional binary black hole gravitational wave candidates.

@@

Until recently, there was no definitive evidence for intermediate-mass black holes with a mass in between the range of stellar-mass and supermassive black holes ($10^2-10^6\ \msun$). 
\et{There have been numerous claims of IMBH detection. The previous sentence is correct that there are no \textit{definitive} detections before GW190521, but the paragraph is misleading by omission. Insert here an exhaustive review of previous IMBH detection claims, noting who made the claim, the mass, and the method.}
This changed with the detection of GW190521, a unique gravitational wave event that led to the formation of a black hole with a mass  $142\ \msun$, the first confirmed direct discovery of an intermediate-mass black hole~\cite{gw190521}. Although this is the first gravitational-wave observation of an IMBH, ground-based gravitational wave detectors are sensitive to gravitational waves from even more massive systems with a total mass of up to $400\ \msun$~\cite{GWTC1}.

\et{The following paragraph tries to solve a problem that doesn't exist: as noted in comments above, the absence of GWs from IMBH is not surprising. Delete or rewrite if you want to emphasise the difficulty in IMBH detection.}
A possible explanation for the absence of intermediate-mass events may be due to their misclassification as short-duration instrumental noise transients known as glitches~\cite{ pycbc_short_duration_transients, pe_with_glitch, blip_glitches}. These glitches can mimic astrophysical signals and hence decrease the significance of true gravitational wave events.

One method to account for glitches while searching for gravitational waves from coalescing compact binaries is by utilising the astrophysical Bayesian odds~\cite{bci, BCR1, BCR2, bcr_gw151216, bayesian_odds}. A true Bayesian odds, calculated without using bootstrap techniques, can provide more accurate estimate of significance that does note depend on the search pipeline~\cite{BCR2, bcr_gw151216,  bayesian_odds}.
In this paper, we utilise a Bayesian method, called the Bayesian Coherence Ratio \bcr~\cite{BCR1}, to rank the candidate gravitational wave signals from high-mass compact binary coalescences (systems with total masses in the range of $55-500\ \msun$) in the detector data recorded during O2. Although the \bcr, utilising bootstrap techniques, does not provide the true Bayesian odds, it provides a more astrophysical measure of candidate events' significance than a traditional match-filter significance.
\et{This pitch, that the method is ``a more astrophysical measure'' is unconvincing. (What does that even mean?) Chat with RS/ET to come up with a better pitch built around the idea that the detection statistic is built on Bayesian inferences results, and that it's a step toward fully Bayesian detection.}

@@

The gravitational wave community uses Bayesian inference to perform parameter estimation and model selection. This work utilises Bayesian inference to calculate the significance of high-mass candidate triggers\footnote{\textit{Trigger:} point in time when the matched-filter SNR is larger than a threshold value for a given gravitational wave template~\cite{pycbc_og1}. \textit{Candidate Trigger:} trigger that may be from astrophysical origins.} in LIGO's second observing run. We calculate the significance of triggers using the Bayesian Coherence Ratio~\cite{BCR1}, \bcr, as a ranking statistic, a step toward building a unified Bayesian framework to search for candidates and estimate their parameters.

@@

Bayes theorem states that the posterior probability distribution $p(\parameters|d,\mathcal{H})$ for data $d$ and a vector of parameters \parameters that describe a model which quantifies a hypothesis $\mathcal{H}$, is given by
\begin{equation}
p(\parameters|d,\mathcal{H}) = \frac{\mathcal{L}(d|\parameters, \mathcal{H}) \ \pi(\parameters | \mathcal{H})}{\mathcal{Z}(d|\mathcal{H})}\ , 
\end{equation}
where $\mathcal{L}(d|\parameters, \mathcal{H})$ is the likelihood of the data given the parameters \parameters and the hypothesis, $\pi(\parameters | \mathcal{H})$ is the prior probability of the parameters, and finally,
\begin{equation}
    \mathcal{Z}(d|\mathcal{H}) = \int\limits_{\parameters} \ \mathcal{L}(d|\parameters,\mathcal{H}) \ \pi(\parameters | \mathcal{H}) \ d\parameters
\end{equation} is the likelihood after marginalising over the parameters \parameters.  To compare two hypotheses $\mathcal{H}_A$ and $\mathcal{H}_B$ with the Bayes theorem one can calculate an odds ratio
\begin{equation}
    \mathcal{O}^A_B = \frac{\mathcal{Z}^A\ \pi(\parameters^A)}{\mathcal{Z}^B\ \pi(\parameters^B)}\ ,
\end{equation}
where $\mathcal{Z}^A$ and $\mathcal{Z}^B$ are the shorthand for the evidences  $\mathcal{Z}(d|\mathcal{H}_A)$ and $\mathcal{Z}(d|\mathcal{H}_B)$. The odds ratio can tell us which of the two hypotheses is more likely. For example, if $\mathcal{O}^A_B >> 1$, then this odds ratio indicates that the $\mathcal{H}_A$ describes the data much better than $\mathcal{H}_B$.

The \bcr is a Bayesian odds ratio like the above, of a coherent signal hypotheses $\mathcal{H}_S$ and an incoherent instrumental feature hypothesis $\mathcal{H}_I$ for a network of $D$ detectors. $\mathcal{H_I}$ states that each detector $i$ has either pure stationary Gaussian noise $\mathcal{H}_N$ or Gaussian noise and an incoherent noise transient (glitch). Taking $\mathcal{H}_G$, $Z^S$, $Z^G_i$ and $Z^N_i$ as the Bayesian evidences (marginalised likelihoods) for $\mathcal{H}_S$, $\mathcal{H}_N$, and $\mathcal{H}_G$, the \bcr is given by
\begin{equation}
\label{eq:bcr}
\bcr = \frac{\alpha Z^S}{\prod\limits^D_{i=1} \ [\beta Z^G_i + (1-\beta)Z^N_i]}\ ,
\end{equation}
where $\alpha$ and $\beta$ are the prior-odds of obtaining a signal or a glitch from a stretch of data. The prior-odds can be defined more explicitly as 
\begin{itemize}
    \item $\alpha=P(\mathcal{H}_S)/P(\mathcal{H}_I)$, the prior-odds for obtaining a coherent signal versus an incoherent instrumental feature.
    \item $\beta=P(\mathcal{H}_G| \mathcal{H}_I)$, the prior odds for obtaining a glitch assuming there is an incoherent instrumental feature.
\end{itemize}

When $\mathcal{H}_S$ and $\mathcal{H}_I$ are precisely described and the correct prior-odds are known, the \bcr is a Bayesian odds-ratio. As an odds-ratio, the \bcr is the optimal discriminator between coherent signals and incoherent instrumental features. However, as the priors-odds are unknown, it is invalid to use the \bcr as an odds-ratio. Instead of interpreting the \bcr as a Bayesian odds-ratio, it can be used as a ranking statistic. Using the \bcr as a ranking statistic we can obtain a frequentist significance of a candidate trigger \bcr measured against a {background trigger}\footnote{\textit{Background Triggers}: triggers that do not contain a coherent gravitational wave signal.} \bcr distribution.

When using the \bcr as a detection statistic, the physical interpretation of the prior-odds is lost. Hence, the prior-odds are empirically tuned to maximise the separation between the \bcr distributions for background triggers (expected to favour the $\mathcal{H}_I$ hypothesis) and {simulated triggers}\footnote{\textit{Simulated Triggers}: artificial triggers manufactured by injecting signals into the detector data.} (expected to favour the $\mathcal{H}_S$ hypothesis). Increasing the separation between the two distributions can improve ability of the \bcr to discriminate candidate events as coherent signals or incoherent instrumental features. The tuning process is described in detail in Appendix~\ref{sec:tuning-prior-odds}.

\subsection{Calculating the Significance of Candidate events}
Candidate trigger \bcr values are either statistically insignificant compared to the background \bcr distribution, implying the candidate is more probable to be an incoherent instrumental feature (the $\mathcal{H}_I$ hypothesis), or statistically significant to the background distribution, indicating the possible presence of an astrophysical signal (the $\mathcal{H}_S$ hypothesis). A false alarm probability with trial factors \fap for the candidate triggers can quantify the significance. The \fap is the probability that a candidate trigger originating from a non-astrophysical source can be is falsely identified as a signal.

To calculate the \fap, each candidate \bcr is considered as a single statistical trail that can occur at a fixed false alarm probability $f$, where $f$ is the probability of observing a background \bcr greater than or equal to the candidate \bcr,
\begin{equation}
    f = \frac{\text{Count of } \bcr' \leq \bcr}{\text{Count of } \bcr'} \ .
\end{equation}
The false alarm probability with trails \fap that the \bcr measurement occurs at least once for $N$ trials $(N > 0)$, where $N$ is the number of candidate triggers is
\begin{equation}
    \fap = 1 - (1-f)^N \ .
\end{equation}

@@

\subsection{Triggers for Analysis}

The LIGO Scientific collaboration operates several search pipelines that scan for gravitational waves from compact binary mergers such as \texttt{GstLAL}, \texttt{MBTA}, \texttt{SPIIR} and \pycbc~\cite{GWTC1}.

The output of \pycbc's search is a list of candidate trigger times and their corresponding \pycbc ranking statistic \pycbcstat. The \pycbcstat statistic is akin to the matched-filter signal-to-noise ratio \snr. However, unlike \snr, \pycbcstat includes some information on the candidate signal's intrinsic and extrinsic properties and other information that feeds into determining if the signal can have astrophysical origins~\cite{pycbc_og6}. The additional physical information incorporated in \pycbcstat makes it a more accurate measure of significance than the standard \snr.

Whenever a local maximum of $\pycbcstat > \snr_\text{T}$, where $\snr_\text{T}$ is some threshold value, the \pycbc search pipeline produces a single-detector trigger associated with the detector and time where the apparent signal in the data has its merger~\cite{pycbc_og6}.

For \pycbc to consider a trigger to be a candidate trigger, a trigger that may be from astrophysical origins, the trigger must be observed between detectors with the same template and a time of arrival difference less than the gravitational-wave travel time~\cite{pycbc_og1}. To test its pipeline's sensitivity, \pycbc also conducts searches for simulated triggers. Finally, to quantify the statistical significance of candidate events, \pycbc artificially constructs background triggers to compare against the candidate events. These background triggers are coherent signal-free events, constructed by applying relative offsets, or time-slides, between the data of different detectors~\cite{pycbc_og6}. The time-slides to generate the background triggers are greater than the gravitational-wave travel time between detectors to ensure that the background triggers are not of astrophysical origins.

Our work demonstrates that the \bcr can be used in the same way as \pycbcstat to measure candidate triggers' statistical significance.  The \bcr can be a powerful ranking statistic as the \bcr incorporates information of not only all possible binary black hole systems that might have merged to produce the trigger but also the various incoherent glitches that might cause a false-detection.

@@

\subsection{Acquisition of triggers}
Advanced LIGO's second observing run O2 lasted $38$ weeks~\cite{GWOSC}. The software package, \pycbc~\cite{pycbc_code}, was used by LIGO to process the O2 data in 22 time-frames (approximately 2 weeks for one time-frame) and found several gravitational wave events and numerous gravitational wave candidates~\cite{pycbc_og0, pycbc_og1, pycbc_og2, pycbc_og3, pycbc_og4, pycbc_og5, pycbc_og6}. Some candidate events were vetoed to be glitches, while others were rejected due to their low significance. The data is divided into these time-frames because the detector's sensitivity does not stay constant throughout the eight-month-long observing period.

@@

For our study, we filter the background, simulated and candidate events to include only high-mass events with masses in the ranges of the parameters presented in Table~\ref{tab:parameters}. A plot of the \pycbc triggers from one time-frame, during April 23 - May 8, 2017, is presented in Figure~\ref{fig:templateBank}. This figure also depicts the gravitational wave templates used during the search through this time-frame of data.

@@

The waveform template we utilise is \imrphenomp, a phenomenological waveform template constructed in the frequency domain that models the in-spiral, merger, and ringdown (IMR) of a compact binary coalescence~\citep{khan2016frequency}. Although there exist gravitational wave templates such as \seob~\cite{seobnrv4phm} which incorporate more physics, such as information on higher-order modes, we use \imrphenomp as it is computationally inexpensive compared to others.

@@

Finally, we neglect detector calibration uncertainty and acquire data from from the Gravitational Wave Open Science Center~\cite{GWOSC}. The data we use is the publicly accessible O2 strain data from the Hanford and Livingston detectors, recorded while the detectors are in ``Science Mode''. We obtain the data using \gwpy~\cite{gwpy}.

@@

Using the tuned prior odds the \bcr for the candidate events can be calculated. Figure~\ref{fig:bcrCdf} shows the \bcr distributions for the background triggers, simulated triggers and candidate events. The bulk of the background and simulated trigger distributions are separate but slightly overlap due to some of the simulated signal's being very faint. The separation suggests that the \bcr can successfully distinguish signals from noise or glitches. The vertical lines in Figure~\ref{fig:bcrCdf} displays the \bcr for gravitational wave candidate events. On comparing the candidate event \bcr values with the background distribution, we can estimate \pastro values for the candidate events.

@@

\subsection{GWTC-1 Events}
All the confirmed gravitational wave events from binary black hole mergers reported in GWTC-1 and within our prior space, (specifically GW170104, GW170608, GW170729, GW170809 and GW170814), have $\pastrobcr$ greater than $0.9$, indicating a high probability of the presence of an astrophysical signal.

In addition to the above confirmed gravitational wave events from GWTC-1, we have also analysed several candidate events from GWTC-1, most of which have low \pastrobcr. For example, consider the candidate event 170412, assigned a $\pastro$ of $0.06$ by \gstlal and has a $\pastrobcr$ of $0.01$. This candidate was reported to be excess power caused due noise appearing non-stationary between 60-200 Hz~\cite{GWTC1}. This candidate acts as an example of how $\pastrobcr$ may be utilised to eliminate candidates originating from terrestrial noise sources.

@@

In this paper, we demonstrate that the Bayesian Coherence Ratio~\cite{BCR1} can be used as a ranking statistic to provide a better measure of significance for gravitational wave candidates by re-analysing the significance of high-mass binary black hole triggers from O2. This method takes a step towards building a unified Bayesian framework that provides a search-pipeline agnostic measure of significance, utilising the same level of physical information incorporated during parameter estimation.

@@

Although our analysis targets high-mass triggers, this method can be extended to include the entire body range of LIGO-detectable gravitational wave sources. Additionally, to further improve the method's infrastructure, we can use more robust gravitational wave templates (such as templates that incorporate higher-order modes), and sophisticated glitch models. Future analysis can also incorporate data from all available detectors in a network to increase the sensitivity of \pastrobcr. The BCR can discern better whether a candidate is a coherent astrophysical candidate or an incoherent glitch with data from more detectors.

\av{as the core of this method is PE, improvements in PE can be adapted into this method}

@@

@@

\subsection{Coherent Signal Model}
We model coherent signal using a binary black hole waveform template \template, where the vector \parameters contains a point in the 15 dimensional space describing precessing binary-black hole mergers. For the signal to be coherent, \parameters must be consistent in each 4-second data segment $d_i$ for a network of $D$ detectors, Hence, the coherent signal evidence is calculated as
\begin{equation}
\label{eq:zs}
Z^S = \int\limits_{\parameters} \prod\limits^{D}_{i=1} \left[ \mathcal{L}(d_i|\template)\right] \pi(\parameters | \mathcal{H}_S)\  \text{d}\parameters \ ,
\end{equation}
where $\pi(\parameters| \mathcal{H}_S)$ is the prior for the parameters in the coherent signal hypothesis, and $\mathcal{L}(d_i|\template))$ is the likelihood for the coherent signal hypothesis that depends on the gravitational wave template \template and its parameters \parameters.

\subsection{Incoherent Glitch Model}
Finally, as glitches are challenging to model and poorly understood, we follow \citet{bci} and utilise a surrogate model for glitches: the glitches are modelled using gravitational wave templates  \template with uncorrelated  parameters amongst the different detectors such that  $\parameters_i \neq \parameters_j$ for two detectors $i$ and $j$~\cite{bci}.  Modelling glitches with \template captures the worst case scenario: when glitches are identical to gravitational wave signals (excluding coherent signals). Thus, we can write $Z^G_i$ as 
\begin{equation}
\label{eq:zg}
Z^G_i = \int\limits_{\parameters} \mathcal{L}(d_i|\template)\ \pi(\parameters| \mathcal{H}_G)\  \text{d}\parameters  \ ,
\end{equation}
where $\pi(\theta| \mathcal{H}_G)$ is the prior for the parameters in the incoherent glitch hypothesis.

>>>>

The most promising avenues to find $100-1000 Msun$ black holes are either to identify stellar binaries, which are expected to be common or to detect them in gravitational radiation. Ground based gravitational-wave limits on the low end of this mass regime are continuing to improve.

@@

Since the 1970s, there has been an accumulation of evidence for stellar-mass and supermassive black holes. In April 2019, the Event Horizon Telescope provided the first visual evidence of the supermassive black hole M87~\cite{m87photo}. As of January 2021, the LIGO Scientific Collaboration has confirmed $\sim50$ binary black hole systems and listed numerous candidate events~\cite{GWTC1, GWTC2, gracedb}. Additionally, since the public release of LIGO's first and second observing run's data (O1 and O2), search pipelines such as \pycbc~\cite{pycbc_code} and a research team at the Institute for Advanced Study (IAS)~\cite{IAS0, IAS1, IAS2}, have independently identified additional binary black hole gravitational-wave candidates.

@@

Until recently, there was no definitive evidence for intermediate-mass black holes with a mass in between the range of stellar-mass and supermassive black holes ($10^2-10^6\ \msun$). 
\et{There have been numerous claims of IMBH detection. The previous sentence is correct that there are no \textit{definitive} detections before GW190521, but the paragraph is misleading by omission. Insert here an exhaustive review of previous IMBH detection claims, noting who made the claim, the mass, and the method.}
This changed with the detection of GW190521, a unique gravitational-wave event that led to the formation of a black hole with a mass  $142\ \msun$, the first confirmed direct discovery of an intermediate-mass black hole~\cite{gw190521}. Although this is the first gravitational-wave observation of an IMBH, ground-based gravitational-wave detectors are sensitive to gravitational-waves from even more massive systems with a total mass of up to $400\ \msun$~\cite{GWTC1}.

\et{The following paragraph tries to solve a problem that doesn't exist: as noted in comments above, the absence of GWs from IMBH is not surprising. Delete or rewrite if you want to emphasise the difficulty in IMBH detection.}
A possible explanation for the absence of intermediate-mass events may be due to their misclassification as short-duration instrumental noise transients known as glitches~\cite{ pycbc_short_duration_transients, pe_with_glitch, blip_glitches}. These glitches can mimic astrophysical signals and hence decrease the significance of true gravitational-wave events.

One method to account for glitches while searching for gravitational-waves from coalescing compact binaries (CBCs) is by utilising the astrophysical Bayesian odds~\cite{bci, BCR1, BCR2, bcr_gw151216, bayesian_odds}. A true Bayesian odds, calculated without using bootstrap techniques, can provide more accurate estimate of significance that does note depend on the search pipeline~\cite{BCR2, bcr_gw151216,  bayesian_odds}.
In this paper, we utilise a Bayesian method, called the Bayesian Coherence Ratio \bcr~\cite{BCR1}, to rank the candidate gravitational-wave signals from high-mass compact binary coalescences (systems with total masses in the range of $55-500\ \msun$) in the detector data recorded during O2. Although the \bcr, utilising bootstrap techniques, does not provide the true Bayesian odds, it provides a more astrophysical measure of candidate events' significance than a traditional match-filter significance.
\et{This pitch, that the method is ``a more astrophysical measure'' is unconvincing. (What does that even mean?) Chat with RS/ET to come up with a better pitch built around the idea that the detection statistic is built on Bayesian inferences results, and that it's a step toward fully Bayesian detection.}

@@

The standard framework to identify CBC gravitational-wave signals hidden in data is by quantifying the significance of candidates with null-hypothesis significance testing. In this framework, the candidates' ranking statistic is compared against a background distribution in a frequentist approach. On the other hand, the standard framework for performing parameter estimation and model selection in gravitational-wave astronomy is Bayesian inference. This work utilises Bayesian inference to calculate the Bayesian Coherence odds-ratio~\cite{BCR1}, \bcr, of high-mass candidates in LIGO's second observing run. We use the \bcr not as an odds-ratio but instead as a ranking statistic, a step toward building a unified Bayesian framework to search for candidates and estimate their parameters.

@@

Bayes theorem states that the posterior probability distribution $p(\parameters|d,\mathcal{H})$ for data $d$ and a vector of parameters \parameters that describe a model which quantifies a hypothesis $\mathcal{H}$, is given by
\begin{equation}
p(\parameters|d,\mathcal{H}) = \frac{\mathcal{L}(d|\parameters, \mathcal{H}) \ \pi(\parameters | \mathcal{H})}{\mathcal{Z}(d|\mathcal{H})}\ , 
\end{equation}
where $\mathcal{L}(d|\parameters, \mathcal{H})$ is the likelihood of the data given the parameters \parameters and the hypothesis, $\pi(\parameters | \mathcal{H})$ is the prior probability of the parameters, and finally,
\begin{equation}
    \mathcal{Z}(d|\mathcal{H}) = \int\limits_{\parameters} \ \mathcal{L}(d|\parameters,\mathcal{H}) \ \pi(\parameters | \mathcal{H}) \ d\parameters
\end{equation} is the likelihood after marginalising over the parameters \parameters.  To compare two hypotheses $\mathcal{H}_A$ and $\mathcal{H}_B$ with the Bayes theorem one can calculate an odds-ratio
\begin{equation}
    \mathcal{O}^A_B = \frac{\mathcal{Z}^A\ \pi(\parameters^A)}{\mathcal{Z}^B\ \pi(\parameters^B)}\ ,
\end{equation}
where $\mathcal{Z}^A$ and $\mathcal{Z}^B$ are the shorthand for the evidences  $\mathcal{Z}(d|\mathcal{H}_A)$ and $\mathcal{Z}(d|\mathcal{H}_B)$. The odds-ratio can tell us which of the two hypotheses is more likely. For example, if $\mathcal{O}^A_B >> 1$, then this odds ratio indicates that the $\mathcal{H}_A$ describes the data much better than $\mathcal{H}_B$.

The \bcr is a Bayesian odds-ratio like the above, of a coherent signal hypotheses $\mathcal{H}_S$ and an incoherent instrumental feature hypothesis $\mathcal{H}_I$ (the null-hypotheses) for a network of $D$ detectors. $\mathcal{H_I}$ states that each detector $i$ has either pure stationary Gaussian noise $\mathcal{H}_N$ or Gaussian noise and an incoherent noise transient (glitch). Taking $\mathcal{H}_G$, $Z^S$, $Z^G_i$ and $Z^N_i$ as the Bayesian evidences (marginalised likelihoods) for $\mathcal{H}_S$, $\mathcal{H}_N$, and $\mathcal{H}_G$, the \bcr is given by
\begin{equation}
\label{eq:bcr}
\bcr = \frac{\alpha Z^S}{\prod\limits^D_{i=1} \ [\beta Z^G_i + (1-\beta)Z^N_i]}\ ,
\end{equation}
where $\alpha$ and $\beta$ are the prior-odds of obtaining a signal or a glitch from a stretch of data. The prior-odds can be defined more explicitly as 
\begin{itemize}
    \item $\alpha=P(\mathcal{H}_S)/P(\mathcal{H}_I)$, the prior-odds for obtaining a coherent signal versus an incoherent instrumental feature.
    \item $\beta=P(\mathcal{H}_G| \mathcal{H}_I)$, the prior-odds for obtaining a glitch assuming there is an incoherent instrumental feature.
\end{itemize}

When $\mathcal{H}_S$ and $\mathcal{H}_I$ are precisely described and the correct prior-odds are known, the \bcr is a Bayesian odds-ratio. As an odds-ratio, the \bcr is the optimal discriminator between coherent signals and incoherent instrumental features. However, as the priors-odds are unknown, it is invalid to use the \bcr as an odds-ratio to make an informed decision about whether a candidate is from an astrophysical or terrestrial source. Instead of interpreting the \bcr as a Bayesian odds-ratio, it can be used as a ranking statistic. Using the \bcr as a ranking statistic we can obtain a frequentist significance of a candidate \bcr-value measured against a background \bcr distribution.

When using the \bcr as a detection statistic, the physical interpretation of the prior-odds is lost. Hence, the prior-odds are empirically tuned to maximise the separation between the \bcr distribution of the background (expected to favour the $\mathcal{H}_I$ hypothesis) and the \bcr distribution of artificially manufactured simulated signals (expected to favour the $\mathcal{H}_S$ hypothesis). Increasing the separation between the two distributions can improve ability of the \bcr to discriminate candidate events as coherent signals or incoherent instrumental features. The tuning process is described in detail in Appendix~\ref{sec:tuning-prior-odds}.

\subsection{Calculating the Significance of Candidates}
Candidate \bcr-values are either statistically insignificant compared to the background \bcr distribution, implying the candidate is more probable to be an incoherent instrumental feature (the $\mathcal{H}_I$ null-hypothesis), or statistically significant to the background distribution, indicating the possible presence of an astrophysical signal (the $\mathcal{H}_S$ hypothesis). A false alarm probability with trial factors, \fap, for the candidate \bcr-values can quantify the significance. The \fap is the probability that a candidate originating from a non-astrophysical source can be is falsely identified as a signal.

To calculate the \fap, each candidate \bcr is considered a single statistical trail that can occur at a fixed false alarm probability $f$, where $f$ is the probability of observing a background \bcr greater than or equal to the candidate \bcr,
\begin{equation}
    f = \frac{\text{Count of } \bcr' \leq \bcr}{\text{Count of } \bcr'} \ .
\end{equation}
The false alarm probability with trails \fap that the \bcr measurement occurs at least once for $N$ trials $(N > 0)$, where $N$ is the number of candidate triggers is
\begin{equation}
    \fap = 1 - (1-f)^N \ .
\end{equation}

@@

\subsection{Data for Analysis}
\avi{Probably need a better section label...}
The LIGO Scientific collaboration operates several search pipelines that scan for gravitational-waves from compact binary mergers such as \texttt{GstLAL}, \texttt{MBTA}, \texttt{SPIIR} and \pycbc~\cite{GWTC1}. The output of \pycbc's search is a list of times and their corresponding \pycbc ranking statistic \pycbcstat values. The \pycbcstat ranking-statistic is akin to the matched-filter signal-to-noise ratio \snr. However, unlike \snr, \pycbcstat includes candidate signal's intrinsic and extrinsic properties and other information that feeds into determining if the signal can have astrophysical origins~\cite{pycbc_og6}. Whenever a local maximum of $\pycbcstat > \snr_\text{T}$, where $\snr_\text{T}$ is some predetermined threshold value, the \pycbc search pipeline produces a single-detector \textit{trigger} associated with the detector and time where the apparent signal in the data has its merger~\cite{pycbc_og6}.

When \pycbc observes a trigger between detectors with coincident parameters and a time of arrival difference less than the gravitational-wave travel time between detectors, the trigger is labelled a \textit{candidate event trigger}, a trigger that may be from astrophysical origins~\cite{pycbc_og1}. To test the pipeline's sensitivity \pycbc also conducts searches for \textit{simulated triggers}, artificial triggers manufactured by injecting signals into the detector data. Finally, to quantify the statistical significance of candidate triggers, \pycbc artificially constructs \textit{background triggers} to compare against the candidate events. These background triggers are coherent signal-free events, constructed by applying relative offsets, or time-slides, between the data of different detectors~\cite{pycbc_og6}. The background trigger's \pycbcstat distribution is used to calculate the candidate trigger's significance,  using null-hypothesis significance testing, under the assumption that all candidate event triggers are due to noise.

Our work demonstrates that the \bcr can be used in the same way as \pycbcstat to measure candidate triggers' statistical significance.  The \bcr can be a powerful ranking statistic as it incorporates information of not only all possible binary black hole systems that might have merged to produce the trigger but also the various incoherent glitches that might cause a false-detection.

@@

\subsection{Acquisition of triggers}
Advanced LIGO's second observing run O2 lasted $38$ weeks~\cite{GWOSC}. The software package, \pycbc~\cite{pycbc_code}, was used by LIGO to process the O2 data in 22 time-frames (approximately 2 weeks for one time-frame) and found several gravitational-wave events and numerous gravitational-wave candidates~\cite{pycbc_og0, pycbc_og1, pycbc_og2, pycbc_og3, pycbc_og4, pycbc_og5, pycbc_og6}. Some candidate events were vetoed to be glitches, while others were rejected due to their low significance. The data is divided into these time-frames because the detector's sensitivity does not stay constant throughout the eight-month-long observing period.

@@

For our study, we filter the background, simulated and candidate events to include only high-mass events with masses in the ranges of the parameters presented in Table~\ref{tab:parameters}. A plot of the \pycbc triggers from one time-frame, during April 23 - May 8, 2017, is presented in Figure~\ref{fig:templateBank}. This figure also depicts the gravitational-wave templates used during the search through this time-frame of data.

@@

The waveform template we utilise is \imrphenomp, a phenomenological waveform template constructed in the frequency domain that models the in-spiral, merger, and ringdown (IMR) of a compact binary coalescence~\citep{khan2016frequency}. Although there exist gravitational-wave templates such as \seob~\cite{seobnrv4phm} which incorporate more physics, such as information on higher-order modes, we use \imrphenomp as it is computationally inexpensive compared to others.

@@

Finally, we neglect detector calibration uncertainty and acquire data from from the gravitational-wave Open Science Center~\cite{GWOSC}. The data we use is the publicly accessible O2 strain data from the Hanford and Livingston detectors, recorded while the detectors are in ``Science Mode''. We obtain the data using \gwpy~\cite{gwpy}.

@@

Using the tuned prior odds the \bcr for the candidate events can be calculated. Figure~\ref{fig:bcrCdf} shows the \bcr distributions for the background triggers, simulated triggers and candidate events. The bulk of the background and simulated trigger distributions are separate but slightly overlap due to some of the simulated signal's being very faint. The separation suggests that the \bcr can successfully distinguish signals from noise or glitches. The vertical lines in Figure~\ref{fig:bcrCdf} displays the \bcr for gravitational-wave candidate events. On comparing the candidate event \bcr values with the background distribution, we can estimate \pastro values for the candidate events.

@@

\subsection{GWTC-1 Events}
All the confirmed gravitational-wave events from binary black hole mergers reported in GWTC-1 and within our prior space, (specifically GW170104, GW170608, GW170729, GW170809 and GW170814), have $\pastrobcr$ greater than $0.9$, indicating a high probability of the presence of an astrophysical signal.

In addition to the above confirmed gravitational-wave events from GWTC-1, we have also analysed several candidate events from GWTC-1, most of which have low \pastrobcr. For example, consider the candidate event 170412, assigned a $\pastro$ of $0.06$ by \gstlal and has a $\pastrobcr$ of $0.01$. This candidate was reported to be excess power caused due noise appearing non-stationary between 60-200 Hz~\cite{GWTC1}. This candidate acts as an example of how $\pastrobcr$ may be utilised to eliminate candidates originating from terrestrial noise sources.

@@

In this paper, we demonstrate that the Bayesian Coherence Ratio~\cite{BCR1} can be used as a ranking statistic to provide a better measure of significance for gravitational-wave candidates by re-analysing the significance of high-mass binary black hole triggers from O2. This method takes a step towards building a unified Bayesian framework that provides a search-pipeline agnostic measure of significance, utilising the same level of physical information incorporated during parameter estimation.

@@

Although our analysis targets high-mass triggers, this method can be extended to include the entire body range of LIGO-detectable gravitational-wave sources. Additionally, to further improve the method's infrastructure, we can use more robust gravitational-wave templates (such as templates that incorporate higher-order modes), and sophisticated glitch models. Future analysis can also incorporate data from all available detectors in a network to increase the sensitivity of \pastrobcr. The BCR can discern better whether a candidate is a coherent astrophysical candidate or an incoherent glitch with data from more detectors.

\avi{as the core of this method is PE, improvements in PE can be adapted into this method}

@@

 We greatly appreciate the contributions of all these computing allocations. All pe performed for this study including test runs and failed simulations used about 2.5M core-hours which under the assumption of 30W per core-hour and a CO2 intensity of electricity of ~ 600 kg CO2 per MWh amounts to a carbon footprint of~ 45t of CO2.

@@

\subsection{Coherent Signal Model}
We model coherent signal using a binary black hole waveform template \template, where the vector \parameters contains a point in the 15 dimensional space describing precessing binary-black hole mergers. For the signal to be coherent, \parameters must be consistent in each 4-second data segment $d_i$ for a network of $D$ detectors, Hence, the coherent signal evidence is calculated as
\begin{equation}
\label{eq:zs}
Z^S = \int\limits_{\parameters} \prod\limits^{D}_{i=1} \left[ \mathcal{L}(d_i|\template)\right] \pi(\parameters | \mathcal{H}_S)\  \text{d}\parameters \ ,
\end{equation}
where $\pi(\parameters| \mathcal{H}_S)$ is the prior for the parameters in the coherent signal hypothesis, and $\mathcal{L}(d_i|\template))$ is the likelihood for the coherent signal hypothesis that depends on the gravitational-wave template \template and its parameters \parameters.

\subsection{Incoherent Glitch Model}
Finally, as glitches are challenging to model and poorly understood, we follow \citet{bci} and utilise a surrogate model for glitches: the glitches are modelled using gravitational-wave templates  \template with uncorrelated  parameters amongst the different detectors such that  $\parameters_i \neq \parameters_j$ for two detectors $i$ and $j$~\cite{bci}.  Modelling glitches with \template captures the worst case scenario: when glitches are identical to gravitational-wave signals (excluding coherent signals). Thus, we can write $Z^G_i$ as 
\begin{equation}
\label{eq:zg}
Z^G_i = \int\limits_{\parameters} \mathcal{L}(d_i|\template)\ \pi(\parameters| \mathcal{H}_G)\  \text{d}\parameters  \ ,
\end{equation}
where $\pi(\theta| \mathcal{H}_G)$ is the prior for the parameters in the incoherent glitch hypothesis.
